Solving \\ the \\ 15-Puzzle \\ \& \\ Rubik's Cube

### 2022-09-10 latex/Chapters/Conclusion.tex
update

\textbf{Learning}
Learnt about puzzles, and also even took the time to learn a couple of sequences of moves that are sufficient to solve \textbf{RC}, 2x2x2 in labout one minute, 3x3x3 in about 2 minutes.
\\
\\
\textbf{Implementation}
Implemented a bunch of solvers. Ideally should do in C++ for speed and multithreaded (esp for MCTS)
\\
\\
\textbf{DRL \& DQL}
Quite impressive, without too much effort, how well they perform. However, I feel to really get more out of them (higher dimensions in particular) would need careful tuning of the convergence criteria.
\\
\\
\textbf{Professional issues}
\\
\\
See \cite{ManyWorlds}
\\
\\
See \cite{https://doi.org/10.1111/1758-5899.12718} and \cite{Bostrom2014}

>>>>

\Section{Learning}
Working on this project has been overall a lot of fun. Even setting aside implementation and computer-experimentation for a moment, I have learnt a great deal about puzzles in general and how to approach thinking about them in a more structured fashion. Specifically about the Rubik's cube, I have even spent the time to learn a few algorithms that are sufficient to solve them by hand. I am now able to solve the 2x2x2 under one minute, 3x3x3 usually around 2 minutes and the 4x4x4 in about 5 minutes. More specifically regarding the techniques discussed in this project, the principles that underly uninformed and informed search, A$^{*}$ and heuristics, multi-stage solvers, and of course the various deep learning methods I have played with (\textbf{DL}, \textbf{DRL} and \textbf{DQL}) all extend to many other puzzle and dimensions and are a great new tool-set to add to my knowledge and experience.

\Section{Implementation}
I am quite satistified to have had the time to implement and play with so many solvers, from handcrafting my own \textbf{SP} \textit{naive} solver, to implementing \textbf{BFS}, \textbf{DFS}, A$^{*}$, admissible heuristics such as Manhattan for the \textbf{SP} and improving on it via linear constraints penalty, and of course all the machinery necessary to fit, train and exploit some \textbf{DL}, \textbf{DRL} and \textbf{DQL} heuristics to guide search algorithms such as  A$^{*}$ and \textbf{MCTS}. Ideally I would have liked to implement things in C++ for speed and multithreading (especially for \textbf{MCTS}) but time was limited so I decided to prioritize breadth and ease of experimentation over depth and efficiency of implementation. The choice of Python definitely made it much easier to integrate with many of the ecosystem's libraries (matplotlib, pandas and pytorch in particular).

\Section{Deep Reinforcement-Q Learning}
Even though the Deep Reinforcement Learning and Deep Q Learning models were quite capricious at time to train. Choosing the convergence mechanisms and criteria, the interplay of generation of new training data with the inner and outer loop of the \textbf{DRL} and \textbf{DQL} algorithms (see \ref{sec:TheoryDLDRL} and \ref{sec:TheoryDQL}) seemed more of an art than a sicence at times). Obviously a lot more tuning, as well as much longer training time and computing resources would have been necessary to make these techniques more competitive at higher dimensions, but overall I remain extremely impressed by how, all things considered - and given how daunting the project seemed at first, it took me little effort to train successful \textbf{DRL} and \textbf{DQL} models capable of solving puzzles of such large state and action spaces.

\Section{Professional issues}
Bug in Kociemba and software design
See \cite{ManyWorlds}
See \cite{https://doi.org/10.1111/1758-5899.12718} and \cite{Bostrom2014}

### 2022-09-10 latex/Chapters/SlidingPuzzle.tex
update

\section{Low dimension}
\label{sec:SPLowDimension}

\subsection{God numbers and hardest puzzles}

@@

\subsection{Manhattan heuristic}
\label{MHComp}
In this section, I verify empirically that, as expected, the overhead of adding penalty in Manhattan++ for the linear constraint (which have all been precomputed and stored in a database) is more than compensated for by the reduction in nodes expansion. I have run my solver script for 2x5 in performance test mode, for both Manhattan and Manhattan++, with 250 randomly shuffled puzzles with \textit{nb\_shuffles} from 0 to 60 by increment of 5, as well as with \textit{nb\_shuffles = $\inf$}. The resulting run time and nodes expansions are as follows:

@@

%-----------------------------------
%	SECTION 2
%-----------------------------------
\section{Intermediary case - 3x3}
\label{sec:S33}

@@

\subsection{Perfect learner}
\label{sec:S33PL}
As discussed in the previous section section \ref{sec:SPLowDimension}, the 3x3 \textbf{SP} is one of the cases I have been able to solve perfectly, since it only has 181,440 possible configurations. Its God number is only 31, which definitely makes it manageable. However, this is already an intermediary size, large enough that it is worth trying and comparing a few different methods, including \textbf{DL}, \textbf{DRL} and \textbf{DQL}. To start with, I ran the PerfectLearner and the results are shown below in figure \ref{fig:33SPPerfectLearning}. It is interesting to note that there are only two hardest configurations (of cost 31) and 221 configurations of cost 30.

\begin{figure}[H]
\centering
\includegraphics[scale=0.42]{./Figures/33SPPerfectLearning}
\caption[33SPPerfectLearning]{Perfect Learning of the 3x3 \textbf{SP}}
\label{fig:33SPPerfectLearning}
\end{figure}

\subsection{Deep reinforcement learner}
\label{sec:S33DRL}
Next I have run the DeepReinforcementLearner with the following main parameters:
\begin{itemize}
\item Update the target network every 1,000 epochs (\textit{update\_target\_network\_frequency=1000}) or when the MSE falls below 10 basis points of the maximum target cost-to-go (\textit{update\_target\_network\_threshold=1e-3})
\item Run for a maximum of 11,000 epochs (\textit{nb\_epochs=11000}) or when the maximum target so far has not been increasing by more than 1\% in the last 5,500 epochs (\textit{max\_target\_uptick=1e-2} and \textit{max\_target\_not\_increasing\_epochs\_pct=0.5})
\item At every update of the target network (\textit{training\_data\_every\_epoch=False}), generate 100 sequences of puzzles (\textit{nb\_sequences=100}), each comprised of a 50-scramble puzzle back to goal state, with all intermediary puzzles along the path (\textit{nb\_shuffles=50}).
\item Use a fully connected network (\textit{network\_type=fully\_connected\_net}) with three hidden layers of size 600, 300, 100 (\textit{layers\_description=(600,300,100)}) and use torch.optim.RMSProp as optimiser (\textit{optimiser=rms\_prop}) with an exponential scheduler (\textit{scheduler=exponential\_scheduler}) with gamma 0.999 (\textit{gamma\_scheduler=0.9999}).
\end{itemize}
Below I show some of the DeepReinforcementLearner's quantities tracked over the epochs (see \ref{fig:33SPDeepReinforcementLearning}). I have shown only the following quantities:
\begin{itemize}
\item learning rate, which decreases due to the scheduler, but gets reset at each network update.
\item max target. One interesting dynamic during learning is that due to the way targets are constructed using value-iteration update, and starting from an all 0 network, the max target only increases by about 1 at best at every network update. That is, the first network only learns how to differentiate goal from non-goal, then the second one learns how to differentiate goals, cost 1 and cost 2+ in some sense. This is why I have introduced the first set of parameters discussed just above. The first one makes sure we do not waste time initially training the left-hand-side network when training is easy (as consist of just goal versus not goal). The second ones make sure that we stop training altogether when the max target does not grow anymore. This is because we do not know a-priori what the God number of a given puzzle/dimension is, but let's assume for the sake of the argument that it is 15. Then that means that after 15 updates of the target network we might not really get so much benefit by keeping training and value-iterating.
\item the MSE loss
\item the MSE loss divided by the max target. The exit criterion \textit{update\_target\_network\_threshold} is based on that quantity, since a loss of e.g. 0.01 means little if the we do not compare it to the possible range that the cost-to-go can take.
\item The percentage of all the possible puzzles of that dimension \textit{seen} by the DeepReinforcementLearner. This is an interesting quantity, as it tells us later on when testing out of sample and comparing with other solvers, how well the solvers are able to generalise. In this case (3x3) we can see that the DeepReinforcementLearner quickly has seen a very large percentage of all puzzles. That being said, it is good to keep in mind that \textit{seen} does not mean really much here, since the learning is unsupervised. This is quite in contrast to the DeepLearner, for which puzzles seen come along with the cost-to-go computed by a \textit{teacher}, usually an optimal or at least efficient solver.
\end{itemize}

@@

\subsection{MCTS DQL}
\label{sec:MTCSHyperParamsEffect}

\begin{figure}[H]
\centering
\includegraphics[align=c, scale=0.6]{./Figures/33SPPerformanceMCTS}
\caption[33SPPerformanceMCTS]{\textbf{MCTS} tuning 3x3 \textbf{SP}}
\label{fig:33SPPerformanceMCTS}
\end{figure}

\subsection{Solvers' comparison}
\label{ssec:33SPSC}
\label{sec:S33DRLDQL}
Let me discuss a comparison of several algorithms on 1000 random puzzles generated for a number of random shuffling (with best-effort-no-backtracking) from 0 to 50 in step of 2, as well as for perfect shuffling (denoted by $\infty$) on the comparison graphs. The results are shown in figure \ref{fig:33SPPerformance}
\\
DL: 100 seq 15 to 31 shuffles 10k epochs connected 600 300 100
\\
Manhattan vs ++
\\
DQL vs DRL
\\
DxL vs Perfect
\\ Put table for nb\_shuffle=inf and highlight best in diff categories (cost, optimality score, run time, nodes)
\\ say how much of the puzzles the solvers have seen

\begin{figure}[H]
\centering
\includegraphics[scale=0.65]{./Figures/33SPPerformance}
%\decoRule
\caption[33SPPerformance]{Solvers' performance comparison 3x3 \textbf{SP}}
\label{fig:33SPPerformance}
\end{figure}

\subsection{Solving the hardest 3x3 problem}
\label{ssec:33SPHard}
To finish with the 3x3 SP, let me try to throw one of the two hardest 3x3 configurations (see subsection \ref{sec:SPLowDimension}) at the different solvers to see how they fare. The results are shown here

@@

%-----------------------------------
%	SECTION 3
%-----------------------------------
\section{3x4}

@@

\section{4x4}

>>>>

\Section{Low dimension}
\label{sec:SPLowDimension}

\Subsection{God numbers and hardest puzzles}

@@

\Subsection{Manhattan heuristic}
\label{MHComp}
In this section, I verify empirically that, as expected, the overhead of adding penalty in Manhattan++ for the linear constraint (which have all been precomputed and stored in a database) is more than compensated for by the reduction in nodes expansion. I have run my solver script for 2x5 in performance test mode, for both Manhattan and Manhattan++, with 250 randomly shuffled puzzles with \textit{nb\_shuffles} from 0 to 60 by increment of 5, as well as with \textit{nb\_shuffles = $\inf$}. The resulting run time and nodes expansions are as follows:

@@

%-----------------------------------
%	SECTION 2
%-----------------------------------
\Section{Intermediary case - 3x3}
\label{sec:S33}

@@

\Subsection{Perfect learner}
\label{sec:S33PL}
As discussed in the previous section section \ref{sec:SPLowDimension}, the 3x3 \textbf{SP} is one of the cases I have been able to solve perfectly, since it only has 181,440 possible configurations. Its God number is only 31, which definitely makes it manageable. However, this is already an intermediary size, large enough that it is worth trying and comparing a few different methods, including \textbf{DL}, \textbf{DRL} and \textbf{DQL}. To start with, I ran the PerfectLearner and the results are shown below in figure \ref{fig:33SPPerfectLearning}. It is interesting to note that there are only two hardest configurations (of cost 31) and 221 configurations of cost 30.

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.4]{./Figures/33SPPerfectLearning}}
  \caption[33SPPerfectLearning]{Perfect Learning of the 3x3 \textbf{SP}}
  \label{fig:33SPPerfectLearning}
\end{figure}

\Subsection{Deep reinforcement learner}
\label{sec:S33DRL}

\begin{comment}

Next I have run the DeepReinforcementLearner with the following main parameters:
\begin{itemize}
\item Update the target network every 1,000 epochs (\textit{update\_target\_network\_frequency=1000}) or when the MSE falls below 10 basis points of the maximum target cost-to-go (\textit{update\_target\_network\_threshold=1e-3})
\item Run for a maximum of 11,000 epochs (\textit{nb\_epochs=11000}) or when the maximum target so far has not been increasing by more than 1\% in the last 5,500 epochs (\textit{max\_target\_uptick=1e-2} and \textit{max\_target\_not\_increasing\_epochs\_pct=0.5})
\item At every update of the target network (\textit{training\_data\_every\_epoch=False}), generate 100 sequences of puzzles (\textit{nb\_sequences=100}), each comprised of a 50-scramble puzzle back to goal state, with all intermediary puzzles along the path (\textit{nb\_shuffles=50}).
\item Use a fully connected network (\textit{network\_type=fully\_connected\_net}) with three hidden layers of size 600, 300, 100 (\textit{layers\_description=(600,300,100)}) and use torch.optim.RMSProp as optimiser (\textit{optimiser=rms\_prop}) with an exponential scheduler (\textit{scheduler=exponential\_scheduler}) with gamma 0.999 (\textit{gamma\_scheduler=0.9999}).
\end{itemize}
Below I show some of the DeepReinforcementLearner's quantities tracked over the epochs (see \ref{fig:33SPDeepReinforcementLearning}). I have shown only the following quantities:
\begin{itemize}
\item learning rate, which decreases due to the scheduler, but gets reset at each network update.
\item max target. One interesting dynamic during learning is that due to the way targets are constructed using value-iteration update, and starting from an all 0 network, the max target only increases by about 1 at best at every network update. That is, the first network only learns how to differentiate goal from non-goal, then the second one learns how to differentiate goals, cost 1 and cost 2+ in some sense. This is why I have introduced the first set of parameters discussed just above. The first one makes sure we do not waste time initially training the left-hand-side network when training is easy (as consist of just goal versus not goal). The second ones make sure that we stop training altogether when the max target does not grow anymore. This is because we do not know a-priori what the God number of a given puzzle/dimension is, but let's assume for the sake of the argument that it is 15. Then that means that after 15 updates of the target network we might not really get so much benefit by keeping training and value-iterating.
\item the MSE loss
\item the MSE loss divided by the max target. The exit criterion \textit{update\_target\_network\_threshold} is based on that quantity, since a loss of e.g. 0.01 means little if the we do not compare it to the possible range that the cost-to-go can take.
\item The percentage of all the possible puzzles of that dimension \textit{seen} by the DeepReinforcementLearner. This is an interesting quantity, as it tells us later on when testing out of sample and comparing with other solvers, how well the solvers are able to generalise. In this case (3x3) we can see that the DeepReinforcementLearner quickly has seen a very large percentage of all puzzles. That being said, it is good to keep in mind that \textit{seen} does not mean really much here, since the learning is unsupervised. This is quite in contrast to the DeepLearner, for which puzzles seen come along with the cost-to-go computed by a \textit{teacher}, usually an optimal or at least efficient solver.
\end{itemize}

@@

\end{comment}

\Subsection{MCTS DQL}
\label{sec:MTCSHyperParamsEffect}
In this subsection I explore the effect on \textbf{MCTS} of the trade-off hyper-parameter $c$ as well as of trimming the resulting tree via \textbf{BFS}. (see the threory section \ref{sec:TheoryMCTS} as well as the implementation section \ref{sec:codesolvers} for more details). To that effect, I have run my Solver performance test with four different values of $c$ (0, 1, 10 and 100), and in each case, with and without \textbf{BFS} trimming. A$^{*}[Manhattan++]$ is also presented to have an optimal solver benchmark. For each level of difficulty, the test present the solvers with the same 150 random configurations (for fairness and to increase the significance of the results). The levels of difficulty (number of scrambles from goal) used here go from 5 to 50 in step of 5. The tests were also run with perfect shuffling (difficulty = $\infty$ on the graphs), corresponding to taking scrambling to $\infty$ as explained in \ref{Theory:SPSSS}). I used a \textit{time\_out} of 5 minutes, which none of the solvers hit on any of the (1,650) configurations they were presented with.
The detailed results (percentage of configurations solved, optimality score, median and max cost, run time and expanded nodes) are shown on the next page \ref{fig:33SPPerformanceMCTS}, but let me make here some remarks on those results:
\begin{itemize}
\item $c=0$ makes no exploration as it simply follows the path of lowest cost-to-go. As $c$ increases, we expect more exploration, and therefore higher run time, more expanded nodes, and higher quality solutions. This is what we observe, where $c=0$ converges (for difficulty=$\infty$) to solutions on average three times longer than the optimal solver but take not much longer to obtain. At the other extreme, $c=100$ takes almost 2 orders of magnitude more time to obtain a solution, but these solutions are only 20-25\% longer than that of the optimal solver
\item trimming via \textbf{BFS} takes hardly any time, and improves solutions' lengths by about 10\% at high values of $c$. At low value of $c$ there is little benefit as expected (since the trees are deep but not bushy at all and have much fewer leaves and branches).
\item The quality of solutions is not amazing, but then given the algorithm has not really any mechanism to guarantee optimality this is not very surprising. 
\item Notice that the number of expanded nodes can be compared within \textbf{MCTS} but given the algorithm is totally different to A$^{*}$ it makes little sense to compare between the different families of solvers. Also one point to note is that \textbf{MCTS} would be fairly easy to multithread (within the same process) using an appropriate language such as C++, and this is what \cite{https://doi.org/10.48550/arxiv.1805.07470} have done, and probably why they get considerably faster and better results.
\item Despite the name \textit{Monte Carlo}, there is no randomness (or even peudo-randomness) in this algorithm, since the paths generated at each iteration are really just a deterministic function of the output of the \textbf{DQL} network (or QHeuristic in my code-jargon).
\end{itemize}

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.6]{./Figures/33SPPerformanceMCTS}}
\caption[33SPPerformanceMCTS]{\textbf{MCTS} tuning 3x3 \textbf{SP}}
\label{fig:33SPPerformanceMCTS}
\end{figure}

\Subsection{Solvers' comparison}
\label{ssec:33SPSC}
\label{sec:S33DRLDQL}
Let me discuss a comparison of several algorithms on 1000 random puzzles generated for a number of random shuffling (with best-effort-no-backtracking) from 0 to 50 in step of 2, as well as for perfect shuffling (denoted by $\infty$) on the comparison graphs. The results are shown in figure \ref{fig:33SPPerformance}
\\
DL: 100 seq 15 to 31 shuffles 10k epochs connected 600 300 100
\\
Manhattan vs ++
\\
DQL vs DRL
\\
DxL vs Perfect
\\ Put table for nb\_shuffle=inf and highlight best in diff categories (cost, optimality score, run time, nodes)
\\ say how much of the puzzles the solvers have seen

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.6]{./Figures/33SPPerformance}}
  \caption[33SPPerformance]{Solvers' performance comparison 3x3 \textbf{SP}}
  \label{fig:33SPPerformance}
\end{figure}

\Subsection{Solving the hardest 3x3 problem}
\label{ssec:33SPHard}
To finish with the 3x3 SP, let me try to throw one of the two hardest 3x3 configurations (see subsection \ref{sec:SPLowDimension}) at the different solvers to see how they fare. The results are shown here

@@

%-----------------------------------
%	SECTION 3
%-----------------------------------
\Section{3x4}

@@

\Section{4x4}
